---
title: "London street dust_initial analysis"
author: "Daniel Schillereff"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


###Install and load packages
```{r}
library(tidyverse)
library(ggpubr)
library(ggplot2)
library(RColorBrewer)
library(gridExtra)
library(reshape2)
library(scales)
library(gghalves)
library(ggdist)
library(dplyr)
library(knitr)    # for kable
library(kableExtra) # for styling the table


```


###Load the XRF data
```{r}
#setwd("C:\\Users\\Danie\\OneDrive - King's College London\\London Street Dust\\XRF results")
setwd("C:\\Users\\User\\OneDrive - King's College London\\London Street Dust\\XRF results")

xrf1 <- read.csv("Londonstreetdust_xrf_25.09.2025.csv")


```

###Group 1, Priority elements: Cd, Cu, Fe, Pb, Zn
```{r Summary table of concentrations}
xrf.1.long <- xrf1 %>% 
                  pivot_longer(cols = "Al":"Zr",
                               cols_vary = "slowest",
                               names_to = c("Element"),
                               values_to = "Concentration", 
                               values_drop_na = TRUE
                                )
Group1.elements <- c("Cd", "Cu", "Fe", "Pb", "Zn")

xrf.Group1.long <- xrf.1.long %>%
  filter(Element %in% Group1.elements) %>%
  filter(!is.na(Concentration))

# Function to calculate geometric mean safely (ignores zeros or negatives)
geo_mean <- function(x) {
  x <- x[x > 0]  # geometric mean only makes sense for positive values
  if(length(x) == 0) return(NA)
  exp(mean(log(x)))
}

#Create a table
summary_table <- xrf.Group1.long %>%
  group_by(Element) %>%
  summarise(
    'Arithmetic Mean' = round(mean(Concentration, na.rm = TRUE), 1),
    'Geometric Mean' = round(geo_mean(Concentration),1),
    Median = round(median(Concentration, na.rm = TRUE), 1),
    SD = round(sd(Concentration, na.rm = TRUE), 1),
    Range = paste0(round(min(Concentration, na.rm = TRUE),0), "–", round(max(Concentration, na.rm = TRUE), 0))
  ) %>% 
    ungroup()

table1 <- summary_table %>%
  kable(
    caption = "Summary statistics of element concentrations (mg/kg)",
    digits = 0,
    align = c("l", "r", "r", "r", "r", "r"),
  ) %>%
  kable_styling(
    full_width = FALSE,
    position = "center",
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  )


```


```{r Boxplots}
xrf.1.long <- xrf1 %>% 
                  pivot_longer(cols = "Al":"Zr",
                               cols_vary = "slowest",
                               names_to = c("Element"),
                               values_to = "Concentration", 
                               values_drop_na = TRUE
                                )
Group1.elements <- c("Cd", "Cu", "Fe", "Pb", "Zn")

xrf.Group1.long <- xrf.1.long %>%
  filter(Element %in% Group1.elements) %>%
  filter(!is.na(Concentration))

xrf.Group1.long %>%
  group_by(Element) %>%
  summarize(n = n(), n_na = sum(is.na(Concentration)), class_type = class(Concentration[1]))

xrf.Group1.long$Concentration <- as.numeric(xrf.Group1.long$Concentration)

#Plotting as horizontal lines
p1 <- ggplot(xrf.Group1.long, aes(x = 1, y = Concentration)) +
  geom_point(
    shape = 95,         # underscore-like horizontal line
    size = 50,           # adjust point size
    alpha = 0.2,        # transparency
    color = "darkgrey"
  ) +
  facet_wrap(~Element, nrow = 1, scales = "free_y") +
  theme_bw() +
  labs(
    x = NULL,
    y = "Concentration (mg/kg)"
  ) +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )


#Now plotting as boxplots
p2 <- ggplot(xrf.Group1.long, aes(x = 1, y = Concentration)) +
  geom_boxplot(
    width = 0.3,        # narrow box
    outlier.shape = NA  # remove default outliers
  ) +
  geom_jitter(
    width = 0.05,       # horizontal jitter
    size = 2,
    alpha = 0.5,
    color = "darkgrey"
  ) +
  facet_wrap(~Element, nrow = 1, scales = "free_y") + # one row, free y-axis per element
  scale_y_continuous(n.breaks = 8, expand = expansion(mult = c(0.05, 0.05))) +  # more tick marks
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank()
  ) +
  labs(
    y = "Concentration (mg/kg)"
  )


```


#Analysing Group 1 data

I've then added as horizontal dashed red lines the continental crustal averages of Rudnick & Gao 2003. The vast majority of samples are heavily enriched in Cu, Pb and Zn compared to crustal (shale) averages. (I need to check, I think Cd was reported in Gao et al. 2008). Fe is generally lower. 

```{r Boxplots with crustal averages plotted}
Wedepohl <- data.frame(
  Element = c("Cd", "Cu", "Fe", "Pb", "Zn"),
  thresh = c(NA, 14.3, 30890, 17, 52)  # example values
)

TaylorMcLennan <- data.frame(
  Element = c("Cd", "Cu", "Fe", "Pb", "Zn"),
  thresh = c(NA, 25, (3.5*10000), 20, 71)  # example values
)

RudnickGao <- data.frame(
  Element = c("Cd", "Cu", "Fe", "Pb", "Zn"),
  thresh = c(NA, 28, (3.525 * 10000), 17, 67)  # example values
)

xrf.1.long <- xrf1 %>% 
                  pivot_longer(cols = "Al":"Zr",
                               cols_vary = "slowest",
                               names_to = c("Element"),
                               values_to = "Concentration", 
                               values_drop_na = TRUE
                                )
Group1.elements <- c("Cd", "Cu", "Fe", "Pb", "Zn")

xrf.Group1.long <- xrf.1.long %>%
  filter(Element %in% Group1.elements) %>%
  filter(!is.na(Concentration))

xrf.Group1.long %>%
  group_by(Element) %>%
  summarize(n = n(), n_na = sum(is.na(Concentration)), class_type = class(Concentration[1]))

xrf.Group1.long$Concentration <- as.numeric(xrf.Group1.long$Concentration)

#Plotting as horizontal lines
p1b <- ggplot(xrf.Group1.long, aes(x = 1, y = Concentration)) +
  geom_point(
    shape = 95,         # underscore-like horizontal line
    size = 50,           # adjust point size
    alpha = 0.2,        # transparency
    color = "darkgrey"
  ) +
   geom_hline(
    data = RudnickGao, 
    aes(yintercept = thresh),
    linetype = "dashed",
    color = "red"
  ) +
  facet_wrap(~Element, nrow = 1, scales = "free_y") +
  theme_bw() +
  labs(
    x = NULL,
    y = "Concentration (mg/kg)"
  ) +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )


#Now plotting as boxplots
p2b <- ggplot(xrf.Group1.long, aes(x = 1, y = Concentration)) +
  geom_boxplot(
    width = 0.3,        # narrow box
    outlier.shape = NA  # remove default outliers
  ) +
  geom_jitter(
    width = 0.05,       # horizontal jitter
    size = 2,
    alpha = 0.5,
    color = "darkgrey"
  ) +
  geom_hline(
    data = RudnickGao, 
    aes(yintercept = thresh),
    linetype = "dashed",
    color = "red"
  ) +
  facet_wrap(~Element, nrow = 1, scales = "free_y") + # one row, free y-axis per element
  scale_y_continuous(n.breaks = 8, expand = expansion(mult = c(0.05, 0.05))) +  # more tick marks
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank()
  ) +
  labs(
    y = "Concentration (mg/kg)"
  )


```



